\proglang{Stan}: A Probabilistic Programming Language

### 2013-12-24 src/docs/papers/stan-language/article.tex
updated with Andrew's and Marcus's comments

We did not set out to build \proglang{Stan} as it currently exists.
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

@@

Because the dense metric performs only a global standardization, the
performance of \proglang{Stan} in models where the relative parameter
scales varies by location, such as hierarchical and latent models, can
suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC)
\citep{GirolamiCalderhead:2011} introduces a location-dependent metric
that can overcome these final hurdles, and a full implementation
within Stan is currently under development.

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.
%
\begin{figure}
\begin{Code}
data { 
  int<lower=0> N; 
  int<lower=0,upper=1> y[N];
} 
parameters {
  real<lower=0,upper=1> theta;
} 
model {
  theta ~ beta(1,1);  // prior
  for (n in 1:N) 
    y[n] ~ bernoulli(theta);  // likelihood
}
\end{Code}
\caption{Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability \code{theta}.  
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta} , although there is no
special behavior for conjugate priors in \proglang{Stan}.  This model
is available in the \proglang{Stan} source distribution in
\code{src/models/basic\_estimators/bernoulli.stan}.

@@

The estimated potential scale reduction statistic $\hat{R}$
(\code{Rhat}) value of 1.0 for the parameter of interest \code{theta}
is consistent with convergence.  \proglang{Stan} uses a more
conservative version of $\hat{R}$ than is usual in packages such as
\pkg{Coda} \citep{PlummerEtAl:2006}, first splitting each chain in half to diagnose
montonically increasing or decreasing samples; \citep{GelmanEtAl:2013}
and \citep{Stan:2013} provide detailed definitions.

@@

\subsection{The ``Matt trick'' for hierarchical models}

Consider the following hierarchical logistic regression model fragment.
%
\begin{Code}
data {
  int<lower=1,upper=K> group[N];  // group of data item n
}  
parameters {
  ...
  vector mu;             // mean coeff
  real<lower=0> sigma;   // coeff scale
  vector[K] beta;        // coeff for group k
}
model {
  beta ~ normal(mu,sigma);
  for (n in 1:N)
    y[n] ~ bernoulli_logit(beta[group[n]] * x[n]);
  ...
}
\end{Code}
%
With the direct parameterization, a change in \code{sigma} is
amplified by the lower-level parameters and introduces a large
change in density across the posterior.  Unfortunately the 
expected density variation of an HMC transition is limited%
%
\footnote{The limited density variation is a ultimately consequence
of a constant metric.  An additional benefit of RMHMC are transitions
that can cover much larger variations in density, making it uniquely
suited to these models.}
%
and many transitions are required to explore the full posterior.
The resulting sampler devolves into a random walk with high autocorrelations.

The following reparameterization employs the ``Matt trick'',
%
\footnote{The reparameterization is named after Matt Hoffman, who
  introduced it during modeling discussions in the \proglang{Stan}
  group meetings.  A similar reparameterization by \citep{Choo:2000}
  is discussed in \citep[Section~5.4.5]{Neal:2011}.}
%
\begin{Code}
parameters {
  vector[K] beta_raw;    
}
transformed parameters {
  vector[K] beta;
  beta <- mu + sigma * beta_raw;
}
model {
  beta ~ normal(0,1);
}
\end{Code}
%
This reparameterization removes the particular correlations amongst the
hierarchical variables that would otherwise limit the effectiveness
of the samplers.

@@

@@

@@

Following the usual convention in mathematics, matrix and array
indexing uses the usual square brackets (\code{[ ]}) operator, and
begins from 1.  \proglang{R}, \proglang{MATLAB}, and \proglang{BUGS},
but not of \proglang{C++}, \proglang{Python}, or \proglang{Java}.
If \code{Sigma} is a matrix, then \code{Sigma[n]} is the \code{n}-th
row of \code{Sigma}.

@@

All probability functions support calculating results up to a constant
proportion, which becomes an additive constant on the log scale.
Constancy here refers to being a numeric literal such as \code{1}, a
constant function such as \code{pi()}, a variable declared as data or
transformed data, or an expression involving only other constants.
Non-constants include parameters, transformed parameters, local
variables declared in the transformed parameters or model statements,
as well as any expression involving a non-constant.

Constant terms are dropped from probability function calculations at
the time the model is compiled, so there is no run-time overhead to
decide which expressions denote constants.%
%
\footnote{Both vectorization and dropping constant terms are
  implemented in \proglang{C++} through template metaprograms that
  infer traits of template arguments to the probability functions.
  Whether to drop constants is configurable through a boolean template
  parameter on the log probability and derivative functions generated
  in \proglang{C++} for a model.}
%
For example, calling \code{normal\_log(y,0,1)}, where \code{y} is a
parameter and \code{0} and \code{1} are constants, does not bother
computing the $\log \sigma$ term in the log density because $\sigma=1$
is a constant.  On the other hand, the $y - 0$ term is calculated and
is not a constant because $y$ is not a constant.

@@

>>>>

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the Stan manual, example models including translations of
  the BUGS examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for Stan's
  interfaces for the command line (CmdStan), Python (PyStan), and
  R (Rstan), and the source code repositories and issue trackers.}

Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

@@

Because the dense metric performs only a global standardization, the
performance of \proglang{Stan} in models where the relative parameter
scales varies by location, such as hierarchical and latent models, can
suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC) introduces
a location-dependent metric that can overcome these final hurdles
\citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
of RHMC based on the metric of \citep{Betancourt:2012}, using a
generalization of NUTS to Riemann manifolds \citep{Betancourt:2013}.

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.
%
\begin{figure}
\begin{Code}
data { 
  int<lower=0> N; 
  int<lower=0,upper=1> y[N];
} 
parameters {
  real<lower=0,upper=1> theta;
} 
model {
  theta ~ beta(1,1);     // prior
  y ~ bernoulli(theta);  // likelihood
}
\end{Code}
\caption{Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in BUGS, although it will run more slowly than the
vectorized form:
%
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.  This model is available in the \proglang{Stan}
source distribution in
\code{src/models/basic\_estimators/bernoulli.stan}.

@@

The estimated potential scale reduction statistic $\hat{R}$
(\code{Rhat}) value of 1.0 for the parameter of interest \code{theta}
is consistent with convergence.  \proglang{Stan} uses a more
conservative version of $\hat{R}$ than is usual in packages such as
\pkg{Coda} \citep{PlummerEtAl:2006}, first splitting each chain in half to diagnose
nonstationary chains; \citep{GelmanEtAl:2013}
and \citep{Stan:2013} provide detailed definitions.

@@

\subsection{The non-centered parameterization for hierarchical models}

Consider the following hierarchical logistic regression model fragment.
%
\begin{Code}
data {
  int<lower=1,upper=K> group[N];  // group of data item n
}  
parameters {
  ...
  vector mu;             // mean coeff
  real<lower=0> sigma;   // coeff scale
  vector[K] beta;        // coeff for group k
}
model {
  beta ~ normal(mu,sigma);
  for (n in 1:N)
    y[n] ~ bernoulli_logit(beta[group[n]] * x[n]);
  ...
}
\end{Code}
%
With this parameterization, the \code{beta[k]} are centered around
\code{mu}.  As a result, a change in \code{sigma} is amplified by the
lower-level parameters and introduces a large change in density across
the posterior.  Unfortunately the expected density variation of an HMC
transition is limited%
%
\footnote{The limited density variation is a ultimately consequence of
  a constant metric.  An additional benefit of RHMC are transitions
  that can cover much larger variations in density, making it uniquely
  suited to these models; see \citep{Neal:2011}.}
%
and many transitions are required to explore the full posterior.  The
resulting sampler devolves into a random walk with high
autocorrelations.

\cite{BetancourtGirolami:2013} discuss for more
analysis of centered vs.\ non-centered parameterizations of
hierarchical models.

The following reparameterization employs a non-centered
parameterization (see \citep{BetancourtGirolami:2013} for an analysis
of the benefits of non-centered parameterizations).
%
\begin{Code}
parameters {
  vector[K] beta_raw;    
}
transformed parameters {
  vector[K] beta;
  beta <- mu + sigma * beta_raw;
}
model {
  beta ~ normal(0,1);
}
\end{Code}
%
This reparameterization removes the particular correlations amongst the
hierarchical variables that would otherwise limit the effectiveness
of the samplers.

@@

Stan follows the \proglang{C++} convention of separating words in
variable names using underbars (\code{\_}), rather than dots
(\code{.}), as used in \proglang{R} and \proglang{BUGS}, or camel case
as used in Java.

@@

In addition to special functions, Stan includes distributions with
alternative parameterizations, such as \code{bernoulli_logit}, which
takes a parameter on the log odds (i.e., logit) scale.  This allows a
more concise notation for generalized linear models as well as more
efficient and arithmetically stable execution.

@@

Following the usual convention in mathematics, matrix and array
indexing uses the usual square brackets (\code{[ ]}) operator, and
begins indexing from 1.  If \code{Sigma} is a matrix, then
\code{Sigma[n]} is the \code{n}-th row of \code{Sigma}.

@@

All probability functions support calculating results up to a constant
proportion, which becomes an additive constant on the log scale.
Constancy here refers to being a numeric literal such as \code{1} or
\code{0.5}, a constant function such as \code{pi()}, data and
transformed data variables, or a function that only depends on
literals, constant functions or data variables.

Non-constants include parameters, transformed parameters, local
variables declared in the transformed parameters or model statements,
as well as any expression involving a non-constant.

Constant terms are dropped from probability function calculations at
the time the model is compiled, so there is no run-time overhead to
decide which expressions denote constants.%
%
\footnote{Both vectorization and dropping constant terms are
  implemented in \proglang{C++} through template metaprograms that
  infer traits of template arguments to the probability functions.
  Whether to drop constants is configurable through a boolean template
  parameter on the log probability and derivative functions generated
  in \proglang{C++} for a model.}
%
For example, executing \code{normal\_log(y,0,sigma)} only evaluates
\code{log(sigma)} if \code{sigma} is a parameter, transformed
parameter, or a local variable in the transformed parameters or model
block; that is, \code{log(sigma)} is not evaluated if \code{sigma} is
constant as defined above.

@@

In addition to performing full Bayesian inference via posterior
sampling, Stan also can perform optimization (i.e., computation of the
posterior mode).  We are currently working on implementing other
optimization-based inference approaches including variational Bayes,
expectation propagation, and and marginal inference using approximate
integration.  All these algorithms require optimization steps.
